---
title: "Figure6A"
author: "Emiliano Pereira"
date: "2025-08-06"
output:
  pdf_document:
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Intro**
Taxonomic diversity loss simulated by increasing water temperature scenarios and comparing semesters to reproduce the results of Figure 6A from the original publication. [Seasonal dynamics of the coastal microbiome and its association with environmental factors](https://doi.org/10.1038/s41597-023-02266-0).

### **1. Set the environment**
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(vegan)
source("../scripts/resources/increasing_temp_simulation.R")
```

### **2. Load data**
`asvs_workable.tsv` contains the rarefied ASV abundance profiles, with samples as rows and ASVs as columns.  
`samo_metadata_workable.tsv` contains the metadata for the samples.  
`date2season2community.tsv` is a table mapping the date, season, and community columns.

```{r}
ABUND <- read_tsv("../data/asvs_workable.tsv.gz", show_col_types = FALSE)
METADATA <- read_tsv("../data/samo_metadata_workable.tsv", show_col_types = FALSE)
DATE2SEASON2COMMUITY <- read_tsv("../data/date2season2community.tsv", show_col_types = FALSE)
```

### **3. Convert ABUND to long format and map metadata**
```{r}
ABUND_long <- ABUND %>%
              as.data.frame() %>%
              mutate(Date = as.Date(Date)) %>%
              pivot_longer(-Date, names_to = "asv_id", values_to = "abund") %>%
              left_join(x = ., 
                        y  = METADATA %>% dplyr::select(Date, sample_name, `Temperature_C`), 
                        by = c("Date")) %>%
              left_join(x = ., 
                        y  = DATE2SEASON2COMMUITY , 
                        by = "Date") %>%
               filter(abund > 0) # it is critical to remove zeros, other wise asv ranges in S1 and S2 are the same.
```
### **4. Compute temperature range for each semester**

```{r}
semesters2range <- ABUND_long %>% 
                        group_by(Community) %>%
                        summarise(min_temp = min(Temperature_C), 
                                  max_temp = max(Temperature_C))

s12range <- semesters2range %>% filter(Community == "S1")
s22range <- semesters2range %>% filter(Community == "S2")
```

### **5. Compute temperature range for each ASV**

```{r}
asvs2range <- ABUND_long %>% 
              group_by(asv_id) %>%
              summarise(min_temp = min(Temperature_C), 
                        max_temp = max(Temperature_C)) %>%
              rename(seq_id = asv_id)
```

### **6. Extract communities associated to each semester**

```{r}
ABUND <- ABUND %>%
         column_to_rownames(var = "Date")

s1_i <- DATE2SEASON2COMMUITY %>% 
        filter(Community == "S1") %>% 
        pull(Date) %>%
        as.character()

ABUND_s1 <- ABUND[s1_i,]

s2_i <- DATE2SEASON2COMMUITY %>% 
        filter(Community == "S2") %>% 
        pull(Date) %>%
        as.character()

ABUND_s2 <- ABUND[s2_i,]

# Sanity check
all(colnames(ABUND_s1) == colnames(ABUND_s2))
```
### **7. Simulate diversity loss in semester S1**

```{r}

temp_increase <- seq(0, 3, 0.2)

s1_temp_sim <- increasing_temp_simulation(abund_wide = ABUND_s1,
                                          abund_wide_ref = ABUND_s2,
                                           seqs2range = asvs2range, 
                                           semester2range = s12range,
                                           temp_increase = temp_increase)
```
### **8. Simulate diversity loss in semester S2**

```{r}
s2_temp_sim <- increasing_temp_simulation(abund_wide = ABUND_s2, 
                                          abund_wide_ref = ABUND_s1,
                                          seqs2range = asvs2range, 
                                          semester2range = s22range,
                                          temp_increase = temp_increase)

```
### **9. Extract diversity values from nested lists**

```{r}
s1_temp_div_df <- map_dfr(s1_temp_sim, ~ {
  tibble(
    temp = .x$temp,
    diversity = .x$div$diversity,
    rich_obs = .x$div$rich_obs,
    rich_chao1 = .x$div$rich_chao1,
    semester = "S1"
  )
})

s2_temp_div_df <- map_dfr(s2_temp_sim, ~ {
  tibble(
    temp = .x$temp,
    diversity = .x$div$diversity,
    rich_obs = .x$div$rich_obs,
    rich_chao1 = .x$div$rich_chao1,
    semester = "S2"
  )
})

temp_div_long_df <- rbind(s1_temp_div_df, s2_temp_div_df) %>%
                    pivot_longer(cols = c("diversity", "rich_obs", "rich_chao1"), 
                                 names_to = "metric", values_to = "value") %>%
                    group_by(temp, semester, metric) %>%
                    summarize(mean = mean(value),
                              sd = sd(value)) 
```

### **10. Identify significant differences in diversity loss** 
```{r}

temp_div_long_df2 <- rbind(s1_temp_div_df, s2_temp_div_df) %>%
                      pivot_longer(cols = c("diversity", "rich_obs", "rich_chao1"), 
                                   names_to = "metric", values_to = "value")
               
format_pvalues <- function(anova_welch) {
  
  p_value <- anova_welch$p.value
  
  if (is.na(anova_welch$p.value) == T) {
    
    return(NA)
  } else {
  
    if (p_value < 0.001) {
      p_value <- "<0.001"
    } else {
      if (p_value < 0.01) {
        p_value <- "<0.01"
      } else {
        if (p_value < 0.05) {
          p_value <- "<0.05"
        } else {
          p_value <- ">0.05"
        }
      }
    }
  }
  return(p_value)
  
}

for (i in temp_increase[-1]) {
  
  temp_div_df_i <- temp_div_long_df2 %>%
                   filter(temp == i)
  
  
  # diversity
  welch_anova_i_div <- temp_div_df_i %>%
                       filter(metric == "diversity") %>%
                        oneway.test(value ~ semester, 
                                    data = .,
                                    var.equal = FALSE)
  
  rows_i <- temp_div_long_df$temp == i & temp_div_long_df$metric == "diversity"
  temp_div_long_df[rows_i, "p_value"] <- welch_anova_i_div$p.value
  temp_div_long_df[rows_i, "p_value_formatted"] <- format_pvalues(welch_anova_i_div)
  
  # richness observed
  welch_anova_i_rich_obs <- temp_div_df_i %>%
                           filter(metric == "rich_obs") %>%
                            oneway.test(value ~ semester, 
                                        data = .,
                                        var.equal = FALSE)
  
  rows_i <- temp_div_long_df$temp == i & temp_div_long_df$metric == "rich_obs"
  temp_div_long_df[rows_i, "p_value"] <- welch_anova_i_rich_obs$p.value
  temp_div_long_df[rows_i, "p_value_formatted"] <- format_pvalues(welch_anova_i_rich_obs)
  
  # richness Chao
  welch_anova_i_rich_chao1 <- temp_div_df_i %>%
                               filter(metric == "rich_chao1") %>%
                                oneway.test(value ~ semester, 
                                            data = .,
                                            var.equal = FALSE)
  
  rows_i <- temp_div_long_df$temp == i & temp_div_long_df$metric == "rich_chao1"
  temp_div_long_df[rows_i, "p_value"] <- welch_anova_i_rich_chao1$p.value
  temp_div_long_df[rows_i, "p_value_formatted"] <- format_pvalues(welch_anova_i_rich_chao1)
  
}
```

### **11. Plot diversity loss**
```{r}
text_size <- 8

plot_labels <- c("diversity" = "Shannon diversity", 
                 "rich_obs" = "Observed richness", 
                 "rich_chao1" = "Chao1 estimator")


temp_div_long_df <- temp_div_long_df %>%
                    mutate(y_max = if_else(metric == "diversity", 102, NA),
                           y_max = if_else(metric == "rich_obs", 107, y_max),
                           y_max = if_else(metric == "rich_chao1", 109, y_max))

div_plot <- ggplot(temp_div_long_df, aes(x = temp, y = mean, color = semester)) +
            geom_line() +
            facet_wrap(~metric, scales = "free_y", labeller = as_labeller(plot_labels)) +
            theme_minimal() +
            xlab("Temperature increment (°C)") +
            ylab("Scaled mean index value") +
            scale_y_continuous(breaks = seq(0, 100, 20)) +
            scale_color_manual(values = c("#b14b34","#292c33"), 
                               labels = c("S1" = "Summer-Autumn", "S2" = "Winter-Spring"),
                               name = "Semester") +
            geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),  linewidth = 0.3,  width = 0.1) +
            theme_bw() +
            theme(
              axis.text.x = element_text(size = text_size-2),
              axis.text.y = element_text(size = text_size-2),
              axis.title.x = element_text(size = text_size +2, margin = unit(c(4,0,0,0), "mm")),
              axis.title.y = element_text(size = text_size +2, margin = unit(c(0,2,0,0), "mm")),
              strip.text = element_text(size = text_size+4),
              strip.background = element_blank(),
              # increase legend text 
              legend.text = element_text(size = text_size),
              legend.title = element_text(size = text_size+4)
            ) +
            geom_text(data = temp_div_long_df %>% filter(p_value_formatted == "<0.001"), 
                      aes(x = temp, y = y_max, label = "***", vjust = 0.7), 
                      size = 2, angle = 90, show.legend  = FALSE) +
            geom_text(data = temp_div_long_df %>% filter(p_value_formatted == "<0.01"), 
                      aes(x = temp, y = y_max, label = "**", vjust = 0.7), 
                      size = 2, angle = 90, show.legend  = FALSE) +
            geom_text(data = temp_div_long_df %>% filter(p_value_formatted == "<0.05"), 
                      aes(x = temp, y = y_max, label = "*", vjust = 0.7), 
                      size = 2, angle = 90, show.legend  = FALSE) 

div_plot 
```

### **12. Print session info**
```{r}
sessionInfo()
```
