---
title: "Figure1B"
author: "Emiliano Pereira"
date: "2025-08-06"
output:
  pdf_document:
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Intro**
Non-Metric Multidimensional Scaling (NMDS) performed on the Bray-Curtis compositional dissimilarity matrices of Operational Protein Units (OPUs), to reproduce the analyses and the Figure 1A from the original publication [Seasonal dynamics of the coastal microbiome and its association with environmental factors](https://doi.org/10.1038/s41597-023-02266-0).

### **1. Set the environment**

```{r packages}
library(tidyverse)
library(vegan)
library(ggpubr)
library(corrgram)
library(lubridate)  
library(grid)       
```

### **2. Load data**

```{r load}
ABUND <- read_tsv("../data/opus_workable.tsv.gz", show_col_types = FALSE)
METADATA <- read_tsv("../data/samo_metadata_workable.tsv", show_col_types = FALSE)
```

### 3. Format abundance table
Because the OPU abundance table is in long format (to speed up import), we need to convert it back to wide format.

```{r to-matrix}
ABUND <- ABUND %>%
  tidyr::pivot_wider(names_from = opu_id, values_from = abundance) %>%
  tibble::column_to_rownames("Date") %>%
  as.matrix()
```

### **3. Run NMDS**
NMDS is performed on Bray–Curtis dissimilarities computed from Hellinger-transformed OPU abundance profiles.

```{r}
ABUND_norm <- vegan::decostand(ABUND, method = "hellinger")
set.seed(123)
ABUND_norm_nmds <- vegan::metaMDS(ABUND_norm, k = 2, try = 50, trymax = 50, threshold = 0.9)
```

### **4. Merge tables to plot**

```{r format}
ABUND_norm_nmds_ext <- ABUND_norm_nmds$points %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Date") %>%
  dplyr::mutate(Date = as.Date(Date)) %>%
  dplyr::left_join(METADATA, by = "Date")

# Factor order for seasons
ABUND_norm_nmds_ext$Season <- factor(ABUND_norm_nmds_ext$Season,
                                        levels = c("Summer", "Autumn", "Winter", "Spring"))
```

### **5. Add centroid coordinates**
The centroid coordinates for each season are calculated to draw segments connecting the centroids.

```{r centroids}
ctd_coords <- ABUND_norm_nmds_ext %>%
  dplyr::mutate(Year = lubridate::year(Date)) %>%
  filter(Year == 2018 | Year == 2019) %>%
  dplyr::group_by(Season) %>%
  dplyr::summarize(mean_nmds1 = mean(MDS1), mean_nmds2 = mean(MDS2), .groups = "drop") %>%
  dplyr::arrange(Season)

ctd_coords$segment_x    <- NA
ctd_coords$segment_y    <- NA
ctd_coords$segment_xend <- NA
ctd_coords$segment_yend <- NA

for (i in 1:(nrow(ctd_coords) - 1)) {
  ctd_coords$segment_x[i]    <- ctd_coords$mean_nmds1[i]
  ctd_coords$segment_y[i]    <- ctd_coords$mean_nmds2[i]
  ctd_coords$segment_xend[i] <- ctd_coords$mean_nmds1[i + 1]
  ctd_coords$segment_yend[i] <- ctd_coords$mean_nmds2[i + 1]
}

i_last <- nrow(ctd_coords)
ctd_coords$segment_x[i_last]    <- ctd_coords$mean_nmds1[i_last]
ctd_coords$segment_y[i_last]    <- ctd_coords$mean_nmds2[i_last]
ctd_coords$segment_xend[i_last] <- ABUND_norm_nmds_ext %>% filter(Season == "Summer" & Year == 2020) %>% summarize(mean_nmds1 = mean(MDS1)) %>% pull(mean_nmds1)
ctd_coords$segment_yend[i_last] <- ABUND_norm_nmds_ext %>% filter(Season == "Summer" & Year == 2020) %>% summarize(mean_nmds2 = mean(MDS2)) %>% pull(mean_nmds2)
```

### **7. Plot: MDS1 vs MDS2**

```{r}
x_title <- "MDS1"; y_title <- "MDS2"
x_max <- max(ABUND_norm_nmds_ext$MDS1) + max(ABUND_norm_nmds_ext$MDS1)/10
x_min <- min(ABUND_norm_nmds_ext$MDS1) - abs(min(ABUND_norm_nmds_ext$MDS1)/10)
y_max <- max(ABUND_norm_nmds_ext$MDS2) + max(ABUND_norm_nmds_ext$MDS2)/10
y_min <- min(ABUND_norm_nmds_ext$MDS2) - abs(min(ABUND_norm_nmds_ext$MDS2)/10)

text_size <- 13
season_colors <- c("#c93f1b", "#98482b", "#154360", "#3c7810")
nmds_plot_margin <- c(0.2, 1, 0.2, 0.2)

nmds_plot_mds1_vs_mds2 <- ggplot(ABUND_norm_nmds_ext, aes(MDS1, MDS2, color = Season), alpha = 0.8) +
  geom_point(alpha = 0.95, size = 3) +
  scale_color_manual(values = season_colors) +
  geom_vline(xintercept = 0, color = "gray40", linetype = "dotted") +
  geom_hline(yintercept = 0, color = "gray40", linetype = "dotted") +
  xlab(x_title) + ylab(y_title) +
  theme_bw() +
  theme(
    axis.title = element_text(size = text_size + 4),
    axis.text  = element_text(size = text_size),
    legend.text  = element_text(size = text_size),
    legend.title = element_text(size = text_size),
    plot.margin  = grid::unit(nmds_plot_margin, "lines")
  ) +
  guides(shape = guide_legend(override.aes = list(size = 3)),
         color = guide_legend(override.aes = list(shape = 19, size = 3))) +
  xlim(c(x_min, x_max)) + ylim(c(y_min, y_max)) +
  annotate("text", label = paste("Stress:", round(ABUND_norm_nmds$stress, 3)),
           x = -1.4, y = 0.85, size = 3) +
  geom_segment(data = ctd_coords,
               aes(x = segment_x, y = segment_y, xend = segment_xend, yend = segment_yend),
               arrow = arrow(length = grid::unit(0.1, "inches")),
               color = "gray10", alpha = 0.7, linewidth = 1) +
  geom_text(data = ctd_coords, aes(x = mean_nmds1, y = mean_nmds2),
            label = c("Summer", "Autumn", "Winter", "Spring"),
            vjust = c(-0.5, 0, 1.4, 0), hjust = c(0.5, 1.2, 0, -0.3),
            fontface = "bold", size = 2, show.legend = FALSE) +
  geom_text(aes(label = Date), vjust = 2, size = 1.5, check_overlap = TRUE, show.legend = FALSE)

```

### **8. Create box plot for MDS1**

```{r boxplots, fig.width=7, fig.height=5}
ABUND_norm_nmds_ext$Season_rev <- factor(ABUND_norm_nmds_ext$Season,
                                            levels = c("Spring", "Winter", "Autumn", "Summer"))

x_title_boxplot_mds1 <- x_title
y_title_boxplot_mds1 <- "Season"
boxplot_margin_pc1 <- c(0.2, 7.5, 0.2, 4.85)

boxplot_mds1_vs_season <- ggplot(ABUND_norm_nmds_ext, aes(x = MDS1, y = Season_rev, fill = Season_rev)) +
  geom_boxplot() +
  scale_fill_manual(values = rev(season_colors), name = "Season") +
  xlab(x_title_boxplot_mds1) + ylab(y_title_boxplot_mds1) +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = text_size + 4, margin = grid::unit(c(4,0,0,0), "mm")),
    axis.title.y = element_text(size = text_size + 1, margin = grid::unit(c(0,2,0,0), "mm")),
    axis.text.x  = element_text(size = text_size),
    axis.text.y  = element_text(size = text_size - 3),
    legend.text  = element_text(size = text_size),
    legend.title = element_text(size = text_size),
    plot.margin  = grid::unit(boxplot_margin_pc1, "lines")
  ) +
  xlim(c(x_min, x_max)) +
  geom_segment(aes(x = 0.7, y = 1.5, xend = 0.7, yend = 3.5),
               color = "gray10", alpha = 0.7, linewidth = 0.1) +
  annotate("text", x = 0.75, y = 2.5, label = "***", size = 2, angle = 90)
```
### **9. Create boxplot for MDS2**

```{r}
x_title_boxplot_mds2 <- "Season"
y_title_boxplot_mds2 <- y_title
boxplot_margin_pc2 <- c(0.2, 0.2, 0.2, 0.2)

boxplot_mds2_vs_season <- ggplot(ABUND_norm_nmds_ext, aes(x = Season_rev, y = MDS2, fill = Season_rev)) +
  geom_boxplot() +
  scale_fill_manual(values = rev(season_colors), name = "Season") +
  xlab(x_title_boxplot_mds2) + ylab(y_title_boxplot_mds2) +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = text_size + 4, margin = grid::unit(c(4,0,0,0), "mm")),
    axis.title.y = element_text(size = text_size + 4, margin = grid::unit(c(0,2,0,0), "mm")),
    axis.text    = element_text(size = text_size),
    legend.text  = element_text(size = text_size),
    legend.title = element_text(size = text_size),
    plot.margin  = grid::unit(boxplot_margin_pc2, "lines")
  ) +
  ylim(c(y_min, y_max))

```

### **10. Merge plots**
```{r}

nmds_plot_and_mds1_boxplot <- ggpubr::ggarrange(
  boxplot_mds2_vs_season + ggpubr::rremove("legend") + ggpubr::rremove("x.title") + ggpubr::rremove("x.text"),
  nmds_plot_mds1_vs_mds2 + ggpubr::rremove("x.title") + ggpubr::rremove("x.text") + ggpubr::rremove("y.text") + ggpubr::rremove("y.title"),
  ncol = 2, nrow = 1, widths = c(0.25, 0.75)
)

nmds_plot_and_mds1_and_mds2_boxplot <- ggpubr::ggarrange(
  nmds_plot_and_mds1_boxplot,
  boxplot_mds1_vs_season + ggpubr::rremove("legend"),
  nrow = 2, ncol = 1, heights = c(0.70, 0.3)
)
```

### **11. Visualize plot**

```{r}
nmds_plot_and_mds1_and_mds2_boxplot
```

### **12. Perform ANOVA Welch test: MDS1**

```{r welch-mds1}
ABUND_norm_nmds_ext$Semester <- ifelse(ABUND_norm_nmds_ext$Season %in% c("Winter","Spring"),
                                          "Winter-Spring", "Autumn-Summer")
anova_welch_mds1 <- oneway.test(MDS1 ~ Semester, data = ABUND_norm_nmds_ext, var.equal = FALSE)
anova_welch_mds1
```

### **13. Perform ANOVA Welch test: MDS2**

```{r welch-mds2}
ABUND_norm_nmds_ext$Semester2 <- ifelse(ABUND_norm_nmds_ext$Season %in% c("Summer","Spring"),
                                           "Summer-Spring", "Autumn-Winter")
anova_welch_mds2 <- oneway.test(MDS2 ~ Semester2, data = ABUND_norm_nmds_ext, var.equal = FALSE)
anova_welch_mds2
```
