---
title: "Figure6C"
author: "Emiliano Pereira"
date: "2025-08-06"
output:
  pdf_document:
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Intro**
Temperature range analysis of taxa between the Summer–Autumn and Winter–Spring semesters, to reproduce the results of Figure 6C from the original publication [Seasonal dynamics of the coastal microbiome and its association with environmental factors](https://doi.org/10.1038/s41597-023-02266-0).

### **1. Set the environment**
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(vegan)
```

### **2. Load data**
`asvs_workable.tsv`: rarefied ASV abundance table (wide format).  
`samo_metadata_workable.tsv`: metadata for SAMO samples.  
```{r}
ABUND <- read_tsv("../data/asvs/asvs_workable.tsv.gz", show_col_types = FALSE)
METADATA <- read_tsv("../data/metadata/samo_metadata_workable.tsv", show_col_types = FALSE)
```
### **3. Convert ABUND to long format and map metadata**
```{r}
ABUND_long <- ABUND %>%
              mutate(Date = as.Date(Date)) %>%
              pivot_longer(-Date, names_to = "asv_id", values_to = "abund") %>%
              left_join(x = ., 
                        y = METADATA %>% dplyr::select(Date, 
                                                       amplicon_sample_name, 
                                                       Community,
                                                       Temperature_C), 
                        by = c("Date")) %>%
              filter(abund > 0) # it is critical to remove zeros, other wise asv ranges in S1 and S2 are the same. 
```

### **4. Compute temperature range for each ASV**
```{r}
asvs2range <- ABUND_long %>%
                   group_by(asv_id) %>%
                   mutate(max_temp = max(Temperature_C), 
                          min_temp = min(Temperature_C)) %>%
                   mutate(asvs_temp_range = max_temp - min_temp)
```
### **5. Create boxplot comparing Semester's temperature vs ASVs optimal temperature**
```{r}
text_size <- 12

range_temp_boxplot <- ggplot(asvs2range, aes(x = Community, 
                                                  y = asvs_temp_range, fill = Community)) +
                      # add a line at y = 1
                      geom_boxplot(alpha = 0.8) +
                      theme_minimal() +
                      xlab("Semester") +
                      ylab("Temperature range (°C)") +
                      scale_fill_manual(values = c("#b14b34","#292c33"),
                                        labels = c("S1" = "Summer-Autumn", "S2" = "Winter-Spring"),
                                        name = "Semester") +
                      scale_x_discrete(labels= c("S1" = "Summer-Autumn", "S2" = "Winter-Spring")) +
                      theme_bw() +
                      theme(
                        axis.text.x = element_text(size = text_size),
                        axis.text.y = element_text(size = text_size),
                        axis.title.x = element_text(size = text_size +2, margin = unit(c(4,0,0,0), "mm")),
                        axis.title.y = element_text(size = text_size +2, margin = unit(c(0,2,0,0), "mm")),
                        strip.text = element_text(size = text_size+4),
                        strip.background = element_blank(),
                        # increase legend text 
                        legend.text = element_text(size = text_size),
                        legend.title = element_text(size = text_size+4)
                      ) +
                      # remoge legend
                      theme(legend.position = "none") 


range_temp_boxplot_wtext <- range_temp_boxplot +
                          # add text to the plot
                          geom_segment(aes(x = 1.2, y = 12.5, xend = 1.8, yend = 12.5), 
                                       color = "gray10", alpha = 0.7, linewidth = 0.1) +
                          annotate("text", x = 1.5, y = 12.8, label = "***", size = 5)

range_temp_boxplot_wtext

```

### **6. Run Welch's ANOVA comparing the asvs_temp_range**
```{r}
wtest_range_temp <- oneway.test(asvs_temp_range ~ Community, 
                                data = asvs2range, var.equal = FALSE) 

wtest_range_temp
```
### **7. Print session info**
```{r}
sessionInfo()
```
