---
title: "Figure6C"
author: "Emiliano Pereira"
date: "2025-08-06"
output:
  pdf_document:
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Intro**
Temperature range analysis of taxa between the Summer–Autumn and Winter–Spring semesters, to reproduce the results of Figure 6D from the original publication [Seasonal dynamics of the coastal microbiome and its association with environmental factors](https://doi.org/10.1038/s41597-023-02266-0).

### **1. Set the environment**
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(vegan)
```

### **2. Load data**
`asvs_workable.tsv`: rarefied ASV abundance table (wide format).  
`samo_metadata_workable.tsv`: metadata for SAMO samples.  
```{r}
ABUND <- read_tsv("../data/asvs/asvs_workable.tsv.gz", show_col_types = FALSE)
METADATA <- read_tsv("../data/metadata/samo_metadata_workable.tsv", show_col_types = FALSE)
```
### **3. Convert ABUND to long format and map metadata**
```{r}

ABUND_long <- ABUND %>%
              mutate(Date = as.Date(Date)) %>%
              pivot_longer(-Date, names_to = "asv_id", values_to = "abund") %>%
              left_join(x = ., 
                        y  = METADATA %>% dplyr::select(Date, 
                                                        amplicon_sample_name, 
                                                        Community,
                                                        Temperature_C), 
                        by = c("Date")) %>%
              filter(abund > 0) # it is critical to remove zeros, other wise asv ranges in S1 and S2 are the same. 
```
### **4. Compute optimal temperature for each ASV**
```{r}
asvs2optimal_temp <- ABUND_long %>%
                     group_by(asv_id) %>%
                     mutate(asvs_optimal_temp = sum(abund*Temperature_C)/sum(abund)) %>%
                     mutate(asvs_optimal_temp_ratio = asvs_optimal_temp/Temperature_C)

```

### **5. Create boxplot comparing Semester's temperature vs ASVs optimal temperature**
```{r}

text_size <- 12

opt_temp_boxplot <- ggplot(asvs2optimal_temp, aes(x = Community, 
                                                  y = asvs_optimal_temp_ratio, fill = Community)) +
                    # add a line at y = 1
                    geom_hline(yintercept = 1, linetype = "dashed", color = "gray30") +
                    geom_boxplot(alpha = 0.8) +
                    theme_minimal() +
                    xlab("Semester") +
                    ylab("Optimal to sample temperature ratio") +
                    scale_fill_manual(values = c("#b14b34","#292c33"),
                                      labels = c("S1" = "Summer-Autumn", "S2" = "Winter-Spring"),
                                      name = "Semester") +
                    scale_x_discrete(labels= c("S1" = "Summer-Autumn", "S2" = "Winter-Spring")) +
                    # set breaks in y axis
      
                    scale_y_continuous(breaks = seq(0.5, 2, 0.5), 
                                       labels = seq(0.5, 2, 0.5), 
                                       limits = c(0.5,2.3)) +
                     
                    theme_bw() +
                    theme(
                      axis.text.x = element_text(size = text_size),
                      axis.text.y = element_text(size = text_size),
                      axis.title.x = element_text(size = text_size +2, margin = unit(c(4,0,0,0), "mm")),
                      axis.title.y = element_text(size = text_size +1, margin = unit(c(0,2,0,0), "mm")),
                      strip.text = element_text(size = text_size+4),
                      strip.background = element_blank(),
                      # increase legend text 
                      legend.text = element_text(size = text_size),
                      legend.title = element_text(size = text_size+4)
                    ) +
                    # remoge legend
                    theme(legend.position = "none") 


opt_temp_boxplot_wtext <- opt_temp_boxplot +
                          # add text to the plot
                          annotate("text", x = 2, y = 2.3, label = "***", size = 4) +
                          annotate("text", x = 1, y = 1.5, label = "***", size = 4)

opt_temp_boxplot_wtext


```

### **6. Run Wilcoxon Signed Rank Sum test: Summer-Autumn lower than 1**
```{r}
asvs_optimal_temp_ratio_summer_autumn <- asvs2optimal_temp %>%
                                       filter(Community == "S1") %>%
                                       pull(asvs_optimal_temp_ratio)

wtest_summer_autumn <- wilcox.test(asvs_optimal_temp_ratio_summer_autumn, 
                                  mu = 1, alternative = "less")

wtest_summer_autumn
```
### **7. Run Wilcoxon Signed Rank Sum test: Winter-Spring greater than 1**
```{r}
asvs_optimal_temp_ratio_winter_spring <- asvs2optimal_temp %>%
                                         filter(Community == "S2") %>%
                                         pull(asvs_optimal_temp_ratio)

wtest_winter_spring <- wilcox.test(asvs_optimal_temp_ratio_winter_spring, 
                                 mu = 1, alternative = "greater")

wtest_winter_spring
```

### **8. Print session info**
```{r}
sessionInfo()
```
