---
title: "Figure6B"
author: "Emiliano Pereira"
date: "2025-08-06"
output:
  pdf_document:
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Intro**
Functional diversity loss simulated by increasing water temperature scenarios and comparing semesters to reproduce the results of Figure 6B from the original publication [Seasonal dynamics of the coastal microbiome and its association with environmental factors](https://doi.org/10.1038/s41597-023-02266-0).

### **1. Set the environment**
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(vegan)
source("../scripts/resources/increasing_temp_simulation.R")
```

### **2. Load data**
`opuss_workable.tsv`: rarefied OPUs abundance table (wide format).  
`samo_metadata_workable.tsv`: metadata for SAMO samples.  

```{r}
ABUND <- read_tsv("../data/opus/opus_workable.tsv.gz", show_col_types = FALSE)
METADATA <- read_tsv("../data/metadata/samo_metadata_workable.tsv", show_col_types = FALSE)
```

### **3. Convert ABUND to long format and map metadata**
```{r}
ABUND_long <- ABUND %>%
              left_join(x = ., 
                        y = METADATA %>% dplyr::select(Date, Community, 
                                                       metagenome_sample_name, 
                                                       Temperature_C), 
                        by = c("Date")) %>%
              filter(abundance > 0) 
```
### **4. Compute temperature range for each semester**

```{r}
semesters2range <- ABUND_long %>% 
                        group_by(Community) %>%
                        summarise(min_temp = min(Temperature_C), 
                                  max_temp = max(Temperature_C))

s12range <- semesters2range %>% filter(Community == "S1")
s22range <- semesters2range %>% filter(Community == "S2")
```
### **5. Compute temperature range for each OPU**

```{r}
opus2range <- ABUND_long %>% 
              group_by(opu_id) %>%
              summarise(min_temp = min(Temperature_C), 
                        max_temp = max(Temperature_C)) %>%
              rename(seq_id = opu_id)
```

### **6. Simulate diversity loss in semester S1**

```{r}

ABUND_wide <- ABUND %>%
              pivot_wider(names_from = opu_id, 
                          values_from = abundance, 
                          values_fill = 0) %>%
              column_to_rownames(var = "Date")

s1_i <- METADATA %>% 
        filter(as.character(Date) %in% rownames(ABUND_wide)) %>%
        filter(Community == "S1") %>% 
        pull(Date) %>%
        as.character()

ABUND_s1 <- ABUND_wide[s1_i,] %>% as.matrix()

s2_i <- METADATA %>% 
        filter(as.character(Date) %in% rownames(ABUND_wide)) %>%
        filter(Community == "S2") %>% 
        pull(Date) %>%
        as.character()

ABUND_s2 <- ABUND_wide[s2_i,] %>% as.matrix()
```

### **7. Simulate diversity loss in semester S1**

```{r}
temp_increase <- seq(0, 3, 0.2)

s1_temp_sim <- increasing_temp_simulation(abund_wide = ABUND_s1,
                                          abund_wide_ref = ABUND_s2,
                                          seqs2range = opus2range, 
                                          semester2range = s12range,
                                          temp_increase = temp_increase)
```

### **8. Simulate diversity loss in semester S2**

```{r}
s2_temp_sim <- increasing_temp_simulation(abund_wide = ABUND_s2, 
                                          abund_wide_ref = ABUND_s1,
                                          seqs2range = opus2range, 
                                          semester2range = s22range,
                                          temp_increase = temp_increase)

```

### **9. Extract diversity values from nested lists**

```{r}
s1_temp_div_df <- map_dfr(s1_temp_sim, ~ {
  tibble(
    temp = .x$temp,
    diversity = .x$div$diversity,
    rich_obs = .x$div$rich_obs,
    rich_chao1 = .x$div$rich_chao1,
    semester = "S1"
  )
})

s2_temp_div_df <- map_dfr(s2_temp_sim, ~ {
  tibble(
    temp = .x$temp,
    diversity = .x$div$diversity,
    rich_obs = .x$div$rich_obs,
    rich_chao1 = .x$div$rich_chao1,
    semester = "S2"
  )
})

temp_div_long_df <- rbind(s1_temp_div_df, s2_temp_div_df) %>%
  pivot_longer(cols = c("diversity", "rich_obs", "rich_chao1"), 
               names_to = "metric", values_to = "value") %>%
  group_by(temp, semester, metric) %>%
  summarize(mean = mean(value),
            sd = sd(value)) 


temp_div_long_df
```

### **10. Identify significant differences in diversity loss** 
```{r}

temp_div_long_df2 <- rbind(s1_temp_div_df, s2_temp_div_df) %>%
  pivot_longer(cols = c("diversity", "rich_obs", "rich_chao1"), 
               names_to = "metric", values_to = "value")

format_pvalues <- function(anova_welch) {
  
  p_value <- anova_welch$p.value
  
  if (is.na(anova_welch$p.value) == T) {
    
    return(NA)
  } else {
    
    if (p_value < 0.001) {
      p_value <- "<0.001"
    } else {
      if (p_value < 0.01) {
        p_value <- "<0.01"
      } else {
        if (p_value < 0.05) {
          p_value <- "<0.05"
        } else {
          p_value <- ">0.05"
        }
      }
    }
  }
  return(p_value)
  
}

for (i in temp_increase[-1]) {
  
  temp_div_df_i <- temp_div_long_df2 %>%
    filter(temp == i)
  
  
  # diversity
  welch_anova_i_div <- temp_div_df_i %>%
    filter(metric == "diversity") %>%
    oneway.test(value ~ semester, 
                data = .,
                var.equal = FALSE)
  
  rows_i <- temp_div_long_df$temp == i & temp_div_long_df$metric == "diversity"
  temp_div_long_df[rows_i, "p_value"] <- welch_anova_i_div$p.value
  temp_div_long_df[rows_i, "p_value_formatted"] <- format_pvalues(welch_anova_i_div)
  
  # richness observed
  welch_anova_i_rich_obs <- temp_div_df_i %>%
    filter(metric == "rich_obs") %>%
    oneway.test(value ~ semester, 
                data = .,
                var.equal = FALSE)
  
  rows_i <- temp_div_long_df$temp == i & temp_div_long_df$metric == "rich_obs"
  temp_div_long_df[rows_i, "p_value"] <- welch_anova_i_rich_obs$p.value
  temp_div_long_df[rows_i, "p_value_formatted"] <- format_pvalues(welch_anova_i_rich_obs)
  
  # richness Chao
  welch_anova_i_rich_chao1 <- temp_div_df_i %>%
    filter(metric == "rich_chao1") %>%
    oneway.test(value ~ semester, 
                data = .,
                var.equal = FALSE)
  
  rows_i <- temp_div_long_df$temp == i & temp_div_long_df$metric == "rich_chao1"
  temp_div_long_df[rows_i, "p_value"] <- welch_anova_i_rich_chao1$p.value
  temp_div_long_df[rows_i, "p_value_formatted"] <- format_pvalues(welch_anova_i_rich_chao1)
  
}

```

### **11. Plot diversity loss** 

```{r}
text_size <- 8

plot_labels <- c("diversity" = "Shannon diversity", 
                 "rich_obs" = "Observed richness", 
                 "rich_chao1" = "Chao1 estimator")

temp_div_long_df <- temp_div_long_df %>%
                    mutate(y_max = if_else(metric == "diversity", 104, NA),
                           y_max = if_else(metric == "rich_obs", 115, y_max),
                           y_max = if_else(metric == "rich_chao1", 120, y_max))

div_plot <- ggplot(temp_div_long_df, aes(x = temp, y = mean, color = semester)) +
            geom_line() +
            facet_wrap(~metric, scales = "free_y", labeller = as_labeller(plot_labels)) +
            theme_minimal() +
            xlab("Temperature increment (°C)") +
            ylab("Scaled mean index value") +
            scale_color_manual(values = c("#b14b34","#292c33"), 
                               labels = c("S1" = "Summer-Autumn", "S2" = "Winter-Spring"),
                               name = "Semester") +
            geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd),  linewidth = 0.3,  width = 0.1) +
            scale_y_continuous(breaks = seq(0, 100, 20)) +
            theme_bw() +
            theme(
              axis.text.x = element_text(size = text_size-2),
              axis.text.y = element_text(size = text_size-2),
              axis.title.x = element_text(size = text_size +2, margin = unit(c(4,0,0,0), "mm")),
              axis.title.y = element_text(size = text_size +2, margin = unit(c(0,2,0,0), "mm")),
              strip.text = element_text(size = text_size+4),
              strip.background = element_blank(),
              # increase legend text 
              legend.text = element_text(size = text_size),
              legend.title = element_text(size = text_size+4)
            )  +
            geom_text(data = temp_div_long_df %>% filter(p_value_formatted == "<0.001"), 
                      aes(x = temp, y = y_max, label = "***", vjust = 0.7), 
                      size = 2, angle = 90, show.legend  = FALSE) +
            geom_text(data = temp_div_long_df %>% filter(p_value_formatted == "<0.01"), 
                      aes(x = temp, y = y_max, label = "**", vjust = 0.7), 
                      size = 2, angle = 90, show.legend  = FALSE) +
            geom_text(data = temp_div_long_df %>% filter(p_value_formatted == "<0.05"), 
                      aes(x = temp, y = y_max, label = "*", vjust = 0.7), 
                      size = 2, angle = 90, show.legend  = FALSE) 

div_plot 

```

### **12. Print session info**
```{r}
sessionInfo()
```
