---
title: "Figure5B"
author: "Emiliano Pereira"
date: "2025-08-06"
output:
  pdf_document:
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Intro**
Compares the bNTI values between semesters to reproduce the results of Figure 5B from the original publication [Seasonal dynamics of the coastal microbiome and its association with environmental factors](https://doi.org/10.1038/s41597-023-02266-0).

### **1. Set the environment**
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(vegan)
```

### **2. Load data**
`bNTI_weighted_26_1000_asvs.csv` weighted beta Nearest Taxin Index (NTI) values for the 1000 most abundant ASVs in the dataset. `samo_metadata_workable.tsv` contains the metadata for the samples.
`date2season2community.tsv` is a table mapping the date, season, and community columns.

```{r}
bNTI <- read_csv("../data/tree_analysis/bNTI_weighted_26_1000_asvs.csv", show_col_types = FALSE,
                  col_names = T) %>% column_to_rownames("...1")
METADATA <- read_tsv("../data/samo_metadata_workable.tsv", show_col_types = FALSE)
DATE2SEASON2COMMUITY <- read_tsv("../data/date2season2community.tsv", show_col_types = FALSE)
```


### **3. Extend metadata table**
```{r}
METADATA_ext <- METADATA %>%
                left_join(DATE2SEASON2COMMUITY, 
                          by =c("Date"))
```

## **4. Convert bNTI to long format**
```{r}
bNTI_long <- bNTI %>% 
             as.data.frame() %>%
             rownames_to_column("sample_name") %>%
             pivot_longer(cols = -sample_name, 
                          names_to = "sample_name_vs", 
                          values_to = "bnti") %>%
             filter(is.na(bnti) != T)

```

### **5. Select samples** 
```{r}

sample2date <- METADATA_ext %>% 
               select(Date, sample_name) 

bNTI_long_sel <- bNTI_long %>%
                 filter(sample_name %in% sample2date$sample_name &
                        sample_name_vs %in% sample2date$sample_name)
```
### **6. Add metadata**
```{r}
METADATA_vars <- METADATA_ext %>% select(sample_name, Date, Community)

bNTI_long_sel_ext <- bNTI_long_sel %>%
                     left_join(METADATA_vars, by = c("sample_name" = "sample_name")) %>%
                     left_join(METADATA_vars, by = c("sample_name_vs" = "sample_name"), suffix = c("","_vs")) %>%
                     mutate(type = paste(Community, Community_vs, sep = "_vs_"),
                            type2 = if_else(Community == Community_vs & Community == "S1","within_S1", NA),
                            type2 = if_else(Community == Community_vs & Community == "S2","within_S2", type2),
                            type2 = if_else(Community != Community_vs,"btween", type2))

```
### **7. Select and order dates from S1 and S2**

```{r}

s1_dates_ord <- METADATA_vars %>%
                filter(Community == "S1") %>%
                select(Date) %>%
                arrange(Date)

s2_dates_ord <- METADATA_vars %>%
                filter(Community == "S2") %>%
                select(Date) %>%
                arrange(Date)

```

### **8. Select consecutive dates from S1** 

```{r}
bNTI_long_sel_ext_s1_list <- list()

for (i in 2:dim(s1_dates_ord)[1]) {
 
  date_i_curr <- s1_dates_ord[i,"Date", drop =T]
  date_i_prev <- s1_dates_ord[i-1,"Date",drop =T]
  dates_i <- c(date_i_curr, date_i_prev)
  dates_i_str <- paste(dates_i[1], dates_i[2])
  
  bNTI_long_sel_ext_s1_list[[dates_i_str]] <- bNTI_long_sel_ext %>%
                                                  filter(as.character(Date) %in% c(dates_i) &
                                                         as.character(Date_vs) %in% c(dates_i))
   
}

# create a data frame from all elements in the list
bNTI_long_sel_ext_s1 <- do.call(rbind, bNTI_long_sel_ext_s1_list)

```

### **9. Select consecutive dates from S2** 
```{r}

bNTI_long_sel_ext_s2_list <- list()

for (i in 2:dim(s2_dates_ord)[1]) {
  
  date_i_curr <- s2_dates_ord[i,"Date", drop =T]
  date_i_prev <- s2_dates_ord[i-1,"Date",drop =T]
  dates_i <- c(date_i_curr, date_i_prev)
  dates_i_str <- paste(dates_i[1], dates_i[2])
  
  bNTI_long_sel_ext_s2_list[[dates_i_str]] <- bNTI_long_sel_ext %>%
                                              filter(as.character(Date) %in% c(dates_i) &
                                                       as.character(Date_vs) %in% c(dates_i))
  
  
}

bNTI_long_sel_ext_s2 <- do.call(rbind, bNTI_long_sel_ext_s2_list)

```

### **10. Concat both data frames**
```{r}
bNTI_long_sel_ext_ord <- rbind(bNTI_long_sel_ext_s1, bNTI_long_sel_ext_s2)
```

#### **11. Get percentage of samples with bNTI greater or lower than 2**

```{r}
bNTI_percs <- bNTI_long_sel_ext_ord %>%
              group_by(type) %>%
              summarise(n = n(), 
                        n_bnti_gt_2 = sum(bnti > 2), 
                        n_bnti_lt_2 = sum(bnti < -2), 
                        perc_bnti_gt_2 = n_bnti_gt_2/n, 
                        perc_bnti_lt_2 = n_bnti_lt_2/n,
                        max_bnti = max(bnti)) %>%
              arrange(desc(n))
```

### **12. Plot bNTI per transition type**

```{r}
text_size <- 10

btni_boxplots <- ggplot(bNTI_long_sel_ext_ord, 
                        aes(x = type, y = bnti, 
                            fill = type)) +
  xlab("Semester") +
  ylab("bNTI") +
  geom_boxplot(alpha = 0.8) +
  geom_hline(yintercept = 2, linetype = "dashed") +
  geom_hline(yintercept = -2, linetype = "dashed") +
  scale_fill_manual(values = c("#b14b34","#292c33")) +
  scale_x_discrete(labels = c("S1_vs_S1" = "Summer-Autumn", 
                              "S2_vs_S2" = "Winter-Spring")) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = text_size, margin = margin(b = 5)),
    axis.text.y = element_text(size = text_size),
    axis.title = element_text(size =  text_size +4),
    strip.text = element_text(size = text_size+4),
    strip.background = element_blank(),
    # extend margin
    plot.margin = margin(1, 1, 1, 1, "cm")
  ) +
  #  add percentage of samples with bnti > 2 above bar plot
  # move the number for better visualization
  geom_text(data = bNTI_percs, aes(x = type, y = 3,
                                   label = scales::percent(perc_bnti_gt_2, accuracy = 0.01),
                                   hjust  = -0.2),
            vjust = -0.5, size = 2.5) +
  guides(fill = "none")

btni_boxplots
```

### **13. Print session info**
```{r}
sessionInfo()
```
